Online Model-Based and Model-Free Reinforcement Learning in Accelerator Operation with Applications to FERMI FEL

### 2020-12-01 main.tex
- paper changers

	\title{Model based and model-free reinforcement learning at FERMI FEL}

@@

	%\tableofcontents
	\section{Introduction and Motivation}
	In particle accelerator operation, one main goal is to provide stable and reproducible performance. In order to achieve this, several control problems have to be considered simultaneously \cite{Kain2020}. Especially if there is no way to model the physics a priori, one might use optimisation techniques as, e.g. derivative-free optimisers (DFOs) \cite{Huang2013,Bruchon2017,Scheinker2020,Hirlaender2019,Welsch2015} or model-based optimisations as Gaussian processes \cite{Hanuka2020,Roussel2020} to restore or maintain the performance. Another promising way is to apply reinforcement learning \cite{Bruchon2020,Bruchon2019,Kain2020,Pang2020} which unveils several advantages over optimisation methods:
	\begin{itemize}
		\item It covers a larger class of problems, as RL optimises a sequence of decisions.
		\item It memorises the problem and does not always begin from zero as a DFO.
		\item Already existing data can be used.
		\item The underlying structure of the problem might be deduced.
	\end{itemize}
	One demanding aspect using RL is the number of iterations needed to train a controller, since RL methods are known to be data-hungry \cite{Sutton2018}, and a second critical aspect is the robustness of the training in real world problems itself.\\
	In this paper, we present the study carried out to solve the maximisation-problem of the radiation intensity generated by a seeded Free-Electron Laser (FEL) on the Free Electron laser Radiation for Multidisciplinary Investigations (FERMI) at Elettra Sincrotrone Trieste.\\
	 Two different algorithm classes were applied successfully to the FERMI FEL problem and reveal different advantages. The critical aspects are addressed, and tailored solutions, generally applicable to problems showing a sample complexity as in accelerator control problems are presented.
	The paper is organised as follows:
	\begin{itemize}
		\item Description of the problem set-up at FERMI
		\item Overview of RL
		\item Design decisions of the implementations used in these studies and theoretical concerns
		\item Results 
		\item Summary and outlook
	\end{itemize}
	% The methods proposed are the iterative Linear Quadratic Regulator (iLQR) - an extension to non-linear systems of the fundamental control theory Linear Quadratic Regulator (LQR) problem -  and the Normalised Advantage Function (NAF) - a continuous variant of the Q-learning algorithm from reinforcement learning field. An exhaustive presentation of both techniques is provided in the following sections.

@@

	The Bellman operator $\bellman^*$ has a unique fixed point but is non-linear, due to the $\max$ - operator:
		\bellman^*\vec Q_\theta(\bs_t,\ba_t) := r(\bs_t,\ba_t)+\gamma\max_\ba\left( Q_\theta(\bs_{t+1},\ba) - Q_\theta(\bs_t,\ba_t)\right).

@@

	\\ Good results were also obtained by following a randomly selected single model each full episode in combination with the  soft-actor-critic (\emph{SAC}), and were used in the experiments labelled \emph{AE-DYNA SAC}. The approach is similar \\

@@

	\section{Discussion and outlook}
	In this paper, the applicability of the method of deep reinforcement learning on the optimization of the FERMI FEL intensity was presented. Two different approaches were tested: a model free and a model based method. Both experiments yielded satisfactory results showing that this non-linear and noisy problem could be solved in a feasibly number of training steps.\\

	\section{Conclusions}
		 The used methods hold tremendous promise for automatizing setups common in accelerators. One purpose of this work was to provide some suggestions how to make advanced deep reinforcement learning techniques work on a real set-up by adjusting available methods.  
	\newpage
	\appendix
	\section{A non-linear standard control problem}
	To provide some transparency of these studies for other labs, we provide results on a famous classical standard control problem \cite{Furutaa}, the \emph{inverted pendulum}.\\
	 It is a non-linear low dimensional unsolved continuous control problem. Unsolved means there is no threshold for the reward to terminate an episode. The episode length, the horizon, is set to 200 steps. In the following several tests were carried out on the \emph{inverted pendulum} to demonstrate the improvements of the selected algorithms, mainly in terms of noise handling. It is of importance when dealing with measurements on real systems.\\
	 For statistical significance, all shown results were performed with five different seeds. The average value and the standard deviation (shaded) are plotted. For this study, we assume that the problem is successfully solved, if the cumulative reward surpasses a threshold of -200. The threshold is indicated by a dashed green line in the corresponding figures.
%	  and a rolling average of 15 steps is used for a better visualization.

@@

	\caption{Cumulative reward of different \emph{NAF} implementations as discussed in the text without noise.}

@@

		\caption{Cumulative reward of different NAF implementations with artificial noise as discussed in the text.}

@@

	\\Using the anchors (the noise level is homeoscetastic) and known in the dynamics model stabilizes the training of the \emph{AE-DYNA} as illustrated in \cref{fig:comparsion_noise_ae_dyna}. It the upper plot the mean cumulative reward on 10 test episodes on the real environment is shown during the training. It should indicate the result if the training is stopped at this training epoch. One cannot observe this during a real training, unless one does costly performance measurements during the training. The blue curve reaches the target much quicker exhibiting less variation. An epoch consists of 3000 iterations of the \emph{SAC}.\\

@@

		\caption{Cumulative reward of \emph{AE-DYNA} with artificial noise using the anchor.}

@@

@@

		\caption{The comparison of the NAF2 and the \emph{AE-DYNA} on the noisy pendulum.}

@@

%	\section{Theoretical aspects on the \emph{AE-DYNA} on FERMI FEL}\label{appendix:Theoretical aspects on the 	AE-DYNA on FERMI FEL}
%	In thist section we want to discuss some theoretical aspects concerning the \emph{AE-DYNA} approach. A simulation was used to obtain the presented analysis. \\
%	One issue of MBRL methods is asymptotic performance. Although good results were obtained using the \emph{AE-DYNA}, there is a difference between the MFRL and MBRL. We call it the \emph{reality gap}. There are ideas to attack the problem, e.g. by learning a meta-policy \cite{Clavera2018} followed by a fine-tuning on the residual physics, which minimises the \emph{reality gap} of training on a simulator and closing than with a small number of training iterations on the real system the gap \cite{Zeng2019}.
%	\begin{figure}[!h]
%		\centering
%		\includegraphics*[width=0.5\textwidth]{Figures/Learning_evolution}
%		\caption{The training .}
%		\label{fig:Learning_evolution}
%	\end{figure}
\section{Still open}
	\begin{itemize}
%	\item Noise benchmarks with pendulum - done
%	\item AE details from papers
%	\item Benchmark NAF2 with pendulum - done
%	\item Compare NAF2 and AE-DYNA
%	\item The first time model based on accelerator problem compared to a novel state of the art MFRL: NAF2
%	\item Tailored to accelerators - short horizons are standard?
	\item Long waiting time in MBRL	
	\item Vary the model size
%	\item Noise sensitivity as mentioned in Benchmark paper
%	\item Improvements of the algorithms:
%	\begin{itemize}
%		\item Adding the noise feature
%		\item Jumping feature using \emph{SAC}
%		\item Noise in the policy
%		\item discuss different approaches to, e.g. dynamic waiting time
%		\item usually prior not discussed in papers
%	\end{itemize}
\end{itemize}

@@

	% \bibliography{bibliography}
	%\addcontentsline{toc}{chapter}{Bibliography}
	%
	%\begin{thebibliography}{9}
	%
	%\bibitem{lsa} D. Jacquet, R. Gorbonosov, G. Kruk, "LSA - The High Level Application Software of the LHC and its Performance during the first 3 years of Operation", ICALEPS, San Francisco, CA, USA, 6 - 11 Oct 2013, pp.thppc058.
	%
	%\bibitem{multiobjective-optimisation} A. Edelen, N. Neveu, M. Frey, Y. Huber, C. Mayes, and A. Adelmann, "Machine learning for orders of magnitude speedup in multiobjective optimisation of particle accelerator systems", Phys. Rev. Accel. Beams 23, 044601, 2020.
	%
	%\bibitem{collimator-alignment} G. Azzopardi, A. Muscat, G. Valentino, S. Redaelli, B. Salvachua, "Operational results of LHC collimator alignment using machine learning", Proc. IPAC'19, Melbourne, Australia, pp. 1208-1211, 2019.
	%
	%\bibitem{SPSAlignment} %InProceedings (Hirlaender2019)
	% S. Hirlaender, M. Fraser, B. Goddard , V. Kain, J. Prieto, L. Stoel, M. Szakaly, F. Velotti, "Automatisation of the SPS ElectroStatic Septa Alignment",  
	%in 10th Int. Partile Accelerator Conf.(IPAC'19), 2019, p. 4001-4004.
	%
	%\bibitem{liu} E. Shaposhnikova et al., "LHC Injectors Upgrade (LIU) Project at CERN", 7$^{th}$ International Particle Accelerator Conference, Busan, Korea, 8 - 13 May 2016, pp.MOPOY059, \url{https://doi.org/10.18429/JACoW-IPAC2016-MOPOY059}.
	%
	%\bibitem{awake} E. Adli, A. Ahuja, O. Apsimon, R. Apsimon, A.-M. Bachmann, D. Barrientos, F. Batsch, J. Bauche, V. B. Olsen, M. Bernardini \textit{et al.}, “Acceleration of electrons in the plasma wakefield of a proton bunch,” Nature 561, 363–367 (2018), \url{https://doi.org/10.1038/s41586-018-0485-4)}.
	%
	%\bibitem{scheinker} A. Scheinker \textit{et al.}, "Online multi-objective particle accelerator optimization of the AWAKE electron beam line for simultaneous emittance and orbit control", AIP Advances 10, 055320 (2020), \url{https://doi.org/10.1063/5.0003423}.
	%
	%\bibitem{barto-sutton} R. Sutton, A. Barto, "Introduction to Reinforcement Learning",  MIT Press, Cambridge, MA, USA, 2018.
	%
	%\bibitem{LQR} H. Kwakernaak, R. Sivan, "Linear Optimal Control Systems", Wiley-Interscience, 1972.
	%
	%\bibitem{SBP} A. Nagabandi, G. Kahn, R. S. Fearing, S. Levine, "Neural Network DYNAmics for Model-Based Deep Reinforcement Learning with Model-Free Fine-Tuning", arXiv:1708.02596, 2017. 
	%
	%\bibitem{DYNA} R. Sutton, "DYNA, an integrated architecture for learning, planning, and reacting", AAAI Spring Symposium, pp. 151-155, 1991, \url{https://doi.org/10.1145/122344.122377}.
	%
	%\bibitem{benchmarkmbrl} T. Wang et al., “Benchmarking Model-Based Reinforcement Learning”,  \url{https://arxiv.org/abs/1907.02057v1}, 2019-07-03.
	%
	%\bibitem{dqn} V. Mnih \textit{et al.}, "Human-level control through deep reinforcement learning", Nature 518, 529–533 (2015).
	%
	%\bibitem{ddpg} T. Lillicrap~\textit{et al.}, "Continuous control with deep reinforcement learning", Proc. ICLR 2016 \url{https://arxiv.org/abs/1509.02971}.
	%
	%\bibitem{naf} S. Gu, T. Lillicrap, I. Sutskever, S. Levine, "Continuous Deep Q-Learning with Model-based Acceleration", in Proc. 33rd International Conference on Machine Learning, New York, NY, USA, 2016.
	%
	%\bibitem{td3} S. Fujimoto, H. van Hoof, D. Meger, "Addressing Function Approximation Error in Actor-Critic Methods", arXiv:1802.09477, 2018.
	%
	%\bibitem{deeppilco} Y. Gal, R. McAllister, and C. E. Rasmussen, “Improving PILCO with Bayesian neural network dynamics models”, in Data-Efficient Machine Learning workshop, ICML, 2016, vol. 4, p. 34.
	%
	%\bibitem{per} T. Schaul, J. Quan, I. Antonoglou, D. Silver, "Prioritised Experience Replay", arXiv:1511.05952, 2015-11-18.
	%
	%\bibitem{pernafgit} S. Hirlaender, \textit{PER-NAF} available at \url{https://github.com/MathPhysSim/PER-NAF/} and
	%\url{https://pypi.org/project/pernaf/}.
	%\bibitem{pyjapc} "pyjapc" available at \url{https://pypi.org/project/pyjapc/}.
	%
	%\bibitem{openai_gym} \url{http://gym.openai.com}
	%
	%\bibitem{madx} MAD-X documentation and source code available at \url{https://mad.web.cern.ch/mad/}.
	%
	%\bibitem{linac4} G. Bellodi, "Linac4 Commissioning Status and Challenges to Nominal Operation", 61$^st$ ICFA Advanced Beam DYNAmics Workshop on High-Intensity and High-Brightness Hadron Beams, Daejeon, Korea, 17 - 22 Jun 2018, pp.MOA1PL03, \url{https://doi.org/10.18429/JACoW-HB2018-MOA1PL03}.
	%
	%\bibitem{auto_matching} F. Velotti, B. Goddard et al., "Automatic AWAKE electron beamline set-up using unsupervised machine learning", to be submitted to Phys. Rev. Accel. Beams.
	%
	%\bibitem{Kumar_1} A. Kumar, A. Gupta, S. Levine, “DisCor: Corrective Feedback in Reinforcement Learning via Distribution Correction,” arXiv:2003.07305, 2020-03-16.
	%
	%\bibitem{Kumar_2} A. Kumar, J. Fu, G. Tucker, S. Levine, “Stabilising Off-Policy Q-Learning via Bootstrapping Error Reduction,” arXiv:1906.00949, 2019-06-03.
	%
	%% NEW REFERENCES
	%\bibitem[bruchon2020basic]{Bruchon, Niky, et al. "Basic Reinforcement Learning Techniques to Control the Intensity of a Seeded Free-Electron Laser." Electronics 9.5 (2020): 781.}
	%% https://www.mdpi.com/2079-9292/9/5/781/htm
	%
	%
	%
	%\end{thebibliography}

>>>>

	\title{Online Model-Based and Model-Free Reinforcement Learning in Accelerator Operation with Applications to FERMI FEL}

@@

	%\tableofcontents
	\section{Introduction and Motivation}
	In particle accelerator operation, one main goal is to provide stable and reproducible performance. In order to achieve this, several control problems have to be considered simultaneously \cite{Kain2020}. Especially if there is no way to model the physics a priori, one might use optimisation techniques as, e.g. derivative-free optimisers (DFOs) \cite{Huang2013,Bruchon2017,Scheinker2020,Hirlaender2019,Welsch2015} or model-based optimisations as Gaussian processes \cite{Hanuka2020,Roussel2020} to restore or maintain the performance. Another promising way is to apply reinforcement learning (RL) \cite{Bruchon2020,Bruchon2019,Kain2020,Pang2020} which unveils several advantages over optimisation methods:
	\begin{itemize}
		\item It covers a larger class of problems, as RL optimises a sequence of decisions.
		\item It memorises the problem and does not always begin from zero as a DFO.
		\item Already existing data can be used.
		\item The underlying structure of the problem might be deduced.
	\end{itemize}
	One demanding aspect using RL is the number of iterations needed to train a controller, since RL methods are known to be data-hungry \cite{Sutton2018}, and a second critical aspect is the robustness of the training in real world problems itself.\\
	In this paper, we present the study carried out to solve the maximisation-problem of the radiation intensity generated by a seeded Free-Electron Laser (FEL) on the Free Electron laser Radiation for Multidisciplinary Investigations (FERMI) at Elettra Sincrotrone Trieste.\\
	 Two different algorithm classes were applied successfully to the FERMI FEL problem and reveal different advantages. The critical aspects are addressed, and tailored solutions, generally applicable to problems showing a sample complexity as in accelerator control problems are presented.
	The paper is organised as follows:
	\begin{itemize}
		\item Description of the problem set-up at FERMI
		\item Overview of RL
		\item Design decisions of the implementations used in these studies and theoretical concerns
		\item Results 
		\item Summary and outlook
	\end{itemize}
	% The methods proposed are the iterative Linear Quadratic Regulator (iLQR) - an extension to non-linear systems of the fundamental control theory Linear Quadratic Regulator (LQR) problem -  and the Normalised Advantage Function (NAF) - a continuous variant of the Q-learning algorithm from reinforcement learning field. An exhaustive presentation of both techniques is provided in the following sections.

@@

	The Bellman operator $\bellman^*$ has a unique fixed point but is non-linear, due to the $\max$ - operator \cite{Sutton2018}:
		\bellman^*\vec Q_\theta(\bs_t,\ba_t) := \\r(\bs_t,\ba_t)+\gamma\max_\ba\left( Q_\theta(\bs_{t+1},\ba) - Q_\theta(\bs_t,\ba_t)\right).

@@

	\\ Good results were also obtained by following a randomly selected single model each full episode in combination with the  soft-actor-critic (\emph{SAC}) \cite{fujimoto2018addressing,Hill2018}, and were used in the experiments labelled \emph{AE-DYNA SAC}. \\

@@

	\section{Discussion and outlook}
	In this paper, the applicability of the method of deep reinforcement learning on the optimization of the FERMI FEL intensity was presented. Two different approaches were tested: a model free and a model based method. Both experiments yielded satisfactory results showing that this non-linear and noisy problem could be solved in a feasibly number of training steps.\\
	The model-free methods were slightly better in the final performance but also more sample were taken. The model-based methods had to be stopped due to operational reasons, hence additional studies regarding a long time performance would be interesting. One big issue when applying model-based methods is the computational effort. In our tests the data acquisition time was only a fraction of the time needed to train the controller on the learned model. In case, the used methods could be parallelized to decrease the computational time.
	\section{Conclusions}
 The used methods hold tremendous promise for automatizing setups common in accelerators. One purpose of this work was to provide some suggestions how to make advanced deep reinforcement learning techniques work on a real set-up by adjusting available methods.\\
 Generally control problems in accelerators have a short horizon, which makes the usage of \emph{DYNA-style} algorithms attractive as the \emph{Me-TRPO} or our  \emph{AE-DYNA} attractive choices for highly sample efficient algorithms.\\

 To provide the possibility for other labs to use the proposed methods the code was released in \url{https://github.com/MathPhysSim/FERMI_RL_Paper} and \cite{Hirlaender2020}.
	\newpage
	\appendix
	\section{A non-linear standard control problem}
	To provide some transparency of these studies for other labs, we provide results on a famous classical standard control problem \cite{Furutaa}, the \emph{inverted pendulum}.\\
	 It is a non-linear low dimensional unsolved continuous control problem. Unsolved means there is no threshold for the reward to terminate an episode. The episode length, the horizon, is set to 200 steps. In the following several tests were carried out on the \emph{inverted pendulum} to demonstrate the improvements of the selected algorithms, mainly in terms of noise handling. It is of importance when dealing with measurements on real systems.\\
	 For statistical significance, all shown results were performed with five different seeds. The average value and the standard deviation (shaded) are plotted. For this study, we assume that the problem is successfully solved, if the cumulative reward surpasses a threshold of -200. The threshold is indicated by a dashed green line in the corresponding figures.
%	  and a rolling average of 15 steps is used for a better visualization.

@@

	\caption{Cumulative reward of different \emph{NAF} implementations as discussed in the text on the \emph{inverted pendulum} without noise.}

@@

		\caption{Cumulative reward of different \emph{NAF} implementations on the \emph{inverted pendulum} with artificial noise as discussed in the text.}

@@

	\\Using the anchors (the noise level is homeoscetastic) and known in the dynamics model stabilizes the training of the \emph{AE-DYNA} as illustrated in \cref{fig:comparsion_noise_ae_dyna}. It the upper plot the mean cumulative reward on 10 test episodes on the real environment is shown during the training. It should indicate the result if the training is stopped at this training epoch. One cannot observe this during a real training, unless one does costly performance measurements during the training. Respecting the aleatoric (\emph{Noise-on-aleatoric}) helps to reach the target much quicker exhibiting less variation compared to the standard use of an ensemble (\emph{Noise-on-non-aleatoric}). An epoch consists of 3000 iterations of the \emph{SAC}.\\

@@

		\caption{Cumulative reward of \emph{AE-DYNA} on the \emph{inverted pendulum} with artificial noise using the anchor.}

@@

\Cref{fig:Compare_models_sizes} shows the impact of the number of models onto the performance on the \emph{inverted pendulum}. The maximum cumulative reward averaged over five different runs tested on the real environment during the training is visualized in dependence of the number of data points. A number of three models shows a good trade-off between performance and
	\includegraphics*[width=0.5\textwidth]{Figures/Compare_models_sizes}
	\caption{Varying the number of models in the ensemble of the \emph{AE-DYNA} on the \emph{inverted pendulum}.}
	\label{fig:Compare_models_sizes}

@@

		\caption{The comparison of the NAF2 and the \emph{AE-DYNA} on the noisy \emph{inverted pendulum}.}

@@

%	\section{Theoretical aspects on the \emph{AE-DYNA} on FERMI FEL}\label{appendix:Theoretical aspects on the 	AE-DYNA on FERMI FEL}
%	In thist section we want to discuss some theoretical aspects concerning the \emph{AE-DYNA} approach. A simulation was used to obtain the presented analysis. \\
%	One issue of MBRL methods is asymptotic performance. Although good results were obtained using the \emph{AE-DYNA}, there is a difference between the MFRL and MBRL. We call it the \emph{reality gap}. There are ideas to attack the problem, e.g. by learning a meta-policy \cite{Clavera2018} followed by a fine-tuning on the residual physics, which minimises the \emph{reality gap} of training on a simulator and closing than with a small number of training iterations on the real system the gap \cite{Zeng2019}.
%	\begin{figure}[!h]
%		\centering
%		\includegraphics*[width=0.5\textwidth]{Figures/Learning_evolution}
%		\caption{The training .}
%		\label{fig:Learning_evolution}
%	\end{figure}
\section{Still open}
	\begin{itemize}
	\item mention the stable baseline.
%	\item Noise benchmarks with pendulum - done
%	\item AE details from papers
%	\item Benchmark NAF2 with pendulum - done
%	\item Compare NAF2 and AE-DYNA
%	\item The first time model based on accelerator problem compared to a novel state of the art MFRL: NAF2
%	\item Tailored to accelerators - short horizons are standard?
%	\item Long waiting time in MBRL	
%	\item Vary the model size
%	\item Noise sensitivity as mentioned in Benchmark paper
%	\item Improvements of the algorithms:
%	\begin{itemize}
%		\item Adding the noise feature
%		\item Jumping feature using \emph{SAC}
%		\item Noise in the policy
%		\item discuss different approaches to, e.g. dynamic waiting time
%		\item usually prior not discussed in papers
%	\end{itemize}
\end{itemize}

@@
